Synthetic prediction methods for minimizing post shock forecasting error

### 2019-11-26 Notes-on-synthetic-prediction.tex
reorganized content; generalized regression result; moved synthetic weighting notes to a separate document

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}

@@

\section{Dynamic panel model}
\label{sec:dpm}
We consider a dynamic panel data model similar with autoregressive structure.  
For simplicity, we will consider the following simplified version with one covariate:
\begin{equation} \label{DPM}
  y_{i,t} = \eta_i + \alpha_iD_{i,t} + \phi_i y_{i,t-1} + \theta_i \x_{i,t} 
    + \varepsilon_{i,t}, 
\end{equation}
where $i = 1,\ldots,J+1$ and $t = 1,\ldots,T$ and $D_{i,t} = 1(t > T_i^*)$, 
$T_i^* < T$.  For simplicity we will assume that $T_i^* = T$ for all $i = 1$,
$\ldots$, $J$.  We will consider a simple random effects structure where 
\begin{align*}
  \eta_i &\overset{iid}{\sim} N(0, \sigma_\eta^2), 
    \qquad i = 1,\ldots J+1, \\
  \alpha_i &\overset{iid}{\sim} N(\mu_\alpha, \sigma_\alpha^2), 
    \qquad i = 1,\ldots J+1, \\
  \phi_i &\overset{iid}{\sim} U(-1,1), 
    \qquad i = 1,\ldots J+1, \\
  \theta_i &\overset{iid}{\sim} U(-1,1), 
    \qquad i = 1,\ldots J+1, \\
  \varepsilon_{i,t} &\overset{iid}{\sim} N(0, \sigma^2), 
    \qquad i = 1,\ldots J+1;\; t = 1, \ldots T, \\
  \eta &\indep \alpha \indep \phi \indep \theta \indep \varepsilon;
\end{align*}
Without loss of generality, we will suppose that we want to forecast for 
$y_{1, T^* + 1}$ where we have no observations for $y_{1,t}, t > T^*$. 
Inference about the random effects parameters is not of interest in this  
forecasting context.  Conditional on all regression parameters 
(except $\alpha$?), previous responses, and covariates, model \eqref{DPM} 
has distribution
$$
  y_{i,t} 
    \sim N(\eta_i + \alpha_iD_{i,t} + \phi_i y_{i,t-1} + \theta_i\x_{i,t}, \sigma^2).
$$
All parameters will be estimated with OLS. In particular, let 
$\hat{\alpha}_i$, $i = 2,\ldots,J+1$ be the OLS estimate of $\alpha_i$ 
and define the adjusted $\alpha$ plugin estimator for time series $i=1$ by,
\begin{equation} \label{adjusted}
  \hat{\alpha}_{\text{adj}} = \frac{1}{J}\sum_{i=2}^{J+1}\hat{\alpha}_i
\end{equation}
where the $\hat{\alpha}_i$s are MLEs of all of the $\alpha_i$s.  
We can use $\hat{\alpha}_{\text{adj}}$ as an estimator for the unknown 
$\alpha_1$ term for which no meaningful estimation information otherwise 
exists.

Consider the candidate forecasts: 
\begin{align*}
  &\text{Forecast 1}: \hat y_{1,T^*+1}^1 = \hat\eta_1 
    + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1}, \\
  &\text{Forecast 2}: \hat y_{1,T^*+1}^2 = \hat\eta_1 
    + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} + \hat{\alpha}_{\text{adj}},
\end{align*}
where $\hat\eta_1$, $\hat\phi_1$, and $\hat\theta_1$ are MLEs of 
$\eta_1$, $\phi_1$, and $\theta_1$ respectively.  
The two forecasts do not differ in their predictions for 
$y_{t}$, $t = 1,\ldots T^*$.  They only differ in predicting $y_{T^*+1}$.  \\

{\bf We want to determine when either $\hat y_{1,T^*+1}^1$ or 
$\hat y_{1,T^*+1}^2$ minimizes the average forecast risk.}

The two forecasts do not differ in their predictions for 
$y_{t}$, $t = 1,\ldots T^*$. \red{repetition.}  They only differ in predicting 
$y_{T^*+1}$.  The forecast risk for $y_{T+1,2}$ is: 
\begin{align*}
  %&R_{T^*+1, 2} - R_{T^*, 2} = 
  &\E(\hat y_{1,T^*+1}^2 - y_{1,T^*+1}^2)^2 \\
  &\qquad= \E\left\{ \hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} 
    + \hat{\alpha}_{\text{adj}}
      - (\eta_1 + \phi_1 y_{1,T^*} + \theta_1 \x_{1,T^*+1} + \alpha_1 + \varepsilon_{1,T^*+1}) \right\}^2 \\
  &\qquad= \E\left\{ (\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1}     
      -\eta_1 - \phi_1 y_{1,T^*} - \theta_1 \x_{1,T^*+1} - \alpha_1 - \varepsilon_{1,T^*+1}) 
        + \hat{\alpha}_{\text{adj}} \right\}^2 \\      
  &\qquad= R_{T, 1} + \E(\hat{\alpha}_{\text{adj}}^2) - 2\E\left\{(\hat\eta_1 + \hat\phi_1 y_{1,T^*} 
    + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} - \theta_1 \x_{1,T^*+1} - \alpha_1 - \varepsilon_{1,T^*+1})
      \hat{\alpha}_{\text{adj}} \right\} \\
  &\qquad= R_{T, 1} + \E(\hat{\alpha}_{\text{adj}}^2) - 2\E(\hat{\alpha}_{\text{adj}})\E(\alpha_1) 
    - 2\E\left\{(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
      - \theta_1 \x_{1,T^*+1})\hat{\alpha}_{\text{adj}} \right\} \\
  &\qquad= R_{T, 1} + \Var(\hat{\alpha}_{\text{adj}}) - \mu_{\alpha}^2 
    - 2\E\left(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
      - \theta_1 \x_{1,T^*+1}\right)\mu_{\alpha},
\end{align*}
where 
$$
  \E(\hat{\alpha}_{\text{adj}}) = \E\left\{\E(\hat{\alpha}_{\text{adj}}| \Hist )\right\}
    = \E\left(\frac{1}{J}\sum_{i=2}^{J+1}\alpha_i\right) 
    = \mu_{\alpha},
$$
and 
$$
  \Hist = \{(\eta_i, \phi_i, \theta_i, \alpha_i, \x_{i,t}, y_{i,t}); 
    i = 2,\ldots,J+1, t = 1,\ldots,T\}.
$$
Define
$$
  \Hist_1 = \{\eta_1, \phi_1, \theta_1, \alpha_1, \x_{1,T^*+1}, (\x_{1,t}, y_{1,t});  
    t = 1,\ldots,T^*\}.
$$
Notice that 
\begin{align*}
  &\E\left(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
    - \theta_1 \x_{1,T^*+1}\right) \\
  &\qquad= \E\left\{\E\left(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
    - \theta_1 \x_{1,T^*+1}\right)|\Hist\cup\Hist_1\right\} = 0,
\end{align*}
(\textbf{is the above true?}) and that 
\begin{align*}
  \Var(\hat{\alpha}_{\text{adj}}) &= \E\left\{\Var(\hat{\alpha}_{\text{adj}}|\Hist)\right\} 
    + \Var\left\{\E(\hat{\alpha}_{\text{adj}}|\Hist)\right\} \\
  &= \E\left\{\Var(\hat{\alpha}_{\text{adj}}|\Hist)\right\} + \frac{\sigma_{\alpha}^2}{J} \\
  &= \frac{1}{J^2} \sum_{i=2}^{J+1} \E\left\{\Var(\hat{\alpha}_j|\Hist)\right\} + \frac{\sigma_{\alpha}^2}{J} \\
  &= \frac{1}{J^2(T-1)}\sum_{i=2}^{J+1} \E\left\{\frac{s_{e,j}^2}{(1-R^2_j)s_{\alpha_j}^2}\right\} 
    + \frac{\sigma_{\alpha}^2}{J},
\end{align*}
where $s_{e,j}^2$ is the estimated variance of model $j$, $R^2_{\alpha_j}$ is 
the multiple $R^2$ obtained from regressing the shock indicator $D$ on the 
other regressors, and $s_j^2$ is... .

@@

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{natbib}

@@

\section{Motivating example: Dynamic panel model}
\label{sec:dpm}

To fix ideas we consider a dynamic panel data model with autoregressive 
structure in \citet{blundell1998initial} without present time period covariate 
information and with a single covariate.  We consider an additional shock 
effect whose presence is given by the binary variable $D_{i,t}$.  Our dynamic 
panel model is  
\begin{equation} \label{DPM}
  y_{i,t} = \eta_i + \alpha_iD_{i,t} + \phi_i y_{i,t-1} + \theta_i \x_{i,t} 
    + \varepsilon_{i,t}, 
\end{equation}
where $i = 1,\ldots,J+1$ and $t = 1,\ldots,T$ and $D_{i,t} = 1(t > T_i^*)$, 
$T_i^* < T$.  For simplicity we will assume that $T_i^* = T$ for all $i = 1$,
$\ldots$, $J$.  We will consider a simple random effects structure where 
\begin{align*}
  \eta_i &\overset{iid}{\sim} N(0, \sigma_\eta^2), 
    \qquad i = 1,\ldots J+1, \\
  \alpha_i &\overset{iid}{\sim} N(\mu_\alpha, \sigma_\alpha^2), 
    \qquad i = 1,\ldots J+1, \\
  \phi_i &\overset{iid}{\sim} U(-1,1), 
    \qquad i = 1,\ldots J+1, \\
  \theta_i &\overset{iid}{\sim} U(-1,1), 
    \qquad i = 1,\ldots J+1, \\
  \varepsilon_{i,t} &\overset{iid}{\sim} N(0, \sigma^2), 
    \qquad i = 1,\ldots J+1;\; t = 1, \ldots T, \\
  \eta &\indep \alpha \indep \phi \indep \theta \indep \varepsilon;
\end{align*}

Our first goal is to evaluate the forecast risk of competing forecasts which 
offer predicts for $y_{1, T^* + 1}$, the response observed after the first 
post shock time period for subject 1.  The difficulty in this forecast stems 
from not observing $\alpha_1$ and not having any readily infromation available 
to estimate it directly.  We show that there is a balance between $\mu_\alpha$ 
and $\sigma_\alpha^2$ that allows us to lower the forecast risk for 
$y_{1, T^* + 1}$ by incorporating what is known about the other $\alpha_i$s 
when such information is available.

Conditional on all regression parameters 
(except $\alpha$?), previous responses, and covariates, model \eqref{DPM} 
has distribution 
$$
  y_{i,t} 
    \sim N(\eta_i + \alpha_iD_{i,t} + \phi_i y_{i,t-1} + \theta_i\x_{i,t}, \sigma^2).
$$
All parameters in this model will be estimated with OLS. In particular, let 
$\hat{\alpha}_i$, $i = 2,\ldots,J+1$ be the OLS estimate of $\alpha_i$ 
and define the adjusted $\alpha$ plugin estimator for time series $i=1$ by,
\begin{equation} \label{adjusted}
  \hat{\alpha}_{\text{adj}} = \frac{1}{J}\sum_{i=2}^{J+1}\hat{\alpha}_i
\end{equation}
where the $\hat{\alpha}_i$s are MLEs of all of the $\alpha_i$s.  
We can use $\hat{\alpha}_{\text{adj}}$ as an estimator for the unknown 
$\alpha_1$ term for which no meaningful estimation information otherwise 
exists.

Consider the candidate forecasts: 
\begin{align*}
  &\text{Forecast 1}: \hat y_{1,T^*+1}^1 = \hat\eta_1 
    + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1}, \\
  &\text{Forecast 2}: \hat y_{1,T^*+1}^2 = \hat\eta_1 
    + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} + \hat{\alpha}_{\text{adj}},
\end{align*}
where $\hat\eta_1$, $\hat\phi_1$, and $\hat\theta_1$ are MLEs of 
$\eta_1$, $\phi_1$, and $\theta_1$ respectively.  
The first forecast ignores the information about the distribution of 
$\alpha_1$ while the second forecast incorporates an estimate of $\mu_\alpha$ 
that is obtained from the other individual forecasts under study.  
Note that the two forecasts do not differ in their predictions for 
$y_{t}$, $t = 1,\ldots T^*$.  They only differ in predicting $y_{T^*+1}$.  \\

We want to determine when either $\hat y_{1,T^*+1}^1$ or $\hat y_{1,T^*+1}^2$ 
minimizes the forecast risk for $\hat y_{1,T^*+1}$.

The forecast risk for $y_{T+1,2}$ is: 
\begin{align*}
  %&R_{T^*+1, 2} - R_{T^*, 2} = 
  &\E(\hat y_{1,T^*+1}^2 - y_{1,T^*+1}^2)^2 \\
  &\qquad= \E\left\{ \hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} 
    + \hat{\alpha}_{\text{adj}}
      - (\eta_1 + \phi_1 y_{1,T^*} + \theta_1 \x_{1,T^*+1} + \alpha_1 + \varepsilon_{1,T^*+1}) \right\}^2 \\
  &\qquad= \E\left\{ (\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1}     
      -\eta_1 - \phi_1 y_{1,T^*} - \theta_1 \x_{1,T^*+1} - \alpha_1 - \varepsilon_{1,T^*+1}) 
        + \hat{\alpha}_{\text{adj}} \right\}^2 \\      
  &\qquad= R_{T, 1} + \E(\hat{\alpha}_{\text{adj}}^2) - 2\E\left\{(\hat\eta_1 + \hat\phi_1 y_{1,T^*} 
    + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} - \theta_1 \x_{1,T^*+1} - \alpha_1 - \varepsilon_{1,T^*+1})
      \hat{\alpha}_{\text{adj}} \right\} \\
  &\qquad= R_{T, 1} + \E(\hat{\alpha}_{\text{adj}}^2) - 2\E(\hat{\alpha}_{\text{adj}})\E(\alpha_1) 
    - 2\E\left\{(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
      - \theta_1 \x_{1,T^*+1})\hat{\alpha}_{\text{adj}} \right\} \\
  &\qquad= R_{T, 1} + \Var(\hat{\alpha}_{\text{adj}}) - \mu_{\alpha}^2 
    - 2\E\left(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
      - \theta_1 \x_{1,T^*+1}\right)\mu_{\alpha},
\end{align*}
where 
$$
  \E(\hat{\alpha}_{\text{adj}}) = \E\left\{\E(\hat{\alpha}_{\text{adj}}| \Hist )\right\}
    = \E\left(\frac{1}{J}\sum_{i=2}^{J+1}\alpha_i\right) 
    = \mu_{\alpha},
$$
and 
$$
  \Hist = \{(\eta_i, \phi_i, \theta_i, \alpha_i, \x_{i,t}, y_{i,t}); 
    i = 2,\ldots,J+1, t = 1,\ldots,T\}.
$$
Define
$$
  \Hist_1 = \{\eta_1, \phi_1, \theta_1, \alpha_1, \x_{1,T^*+1}, (\x_{1,t}, y_{1,t});  
    t = 1,\ldots,T^*\}.
$$
Notice that 
\begin{align*}
  &\E\left(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
    - \theta_1 \x_{1,T^*+1}\right) \\
  &\qquad= \E\left\{\E\left(\hat\eta_1 + \hat\phi_1 y_{1,T^*} + \hat\theta_1 \x_{1,T^*+1} -\eta_1 - \phi_1 y_{1,T^*} 
    - \theta_1 \x_{1,T^*+1}\right)|\Hist\cup\Hist_1\right\} = 0,
\end{align*}
(\textbf{is the above true?}) and that 
\begin{align*}
  \Var(\hat{\alpha}_{\text{adj}}) &= \E\left\{\Var(\hat{\alpha}_{\text{adj}}|\Hist)\right\} 
    + \Var\left\{\E(\hat{\alpha}_{\text{adj}}|\Hist)\right\} \\
  &= \E\left\{\Var(\hat{\alpha}_{\text{adj}}|\Hist)\right\} + \frac{\sigma_{\alpha}^2}{J} \\
  &= \frac{1}{J^2} \sum_{i=2}^{J+1} \E\left\{\Var(\hat{\alpha}_j|\Hist)\right\} + \frac{\sigma_{\alpha}^2}{J} \\
  &= \frac{1}{J^2(T-1)}\sum_{i=2}^{J+1} \E\left\{\frac{s_{e,j}^2}{(1-R^2_j)s_{\alpha_j}^2}\right\} 
    + \frac{\sigma_{\alpha}^2}{J},
\end{align*}
where $s_{e,j}^2$ is the estimated variance of model $j$, $R^2_{\alpha_j}$ is 
the multiple $R^2$ obtained from regressing the shock indicator $D$ on the 
other regressors, and $s_j^2$ is... .

@@

\bibliographystyle{plainnat}
\bibliography{synthetic-prediction-notes}
